The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-01 src/ms.tex
Rview comments addresssed

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface.
This maturity and stability allows for the broader astronomy \python community
to rely upon the \astropypkg core package and build specialized packages within
the Astropy ecosystem.
Even many non-astronomical \python libraries have come to rely on \astropypkg.

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

\subsubsection*{New Features} \label{sec:core-features-new}

\begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
  \item ERFA module is now it’s own thing; Obliquely tie in that units
    remains in astropy
  \item ? Pytest plugin ecosystem
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
  \item WCS API : A shared Python interface for World Coordinate Systems
        Include APE reference
\end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the four main features that entered the astropy core, which we
describe below.

\paragraph{Uncertainties and Distributions.} Early in our carreers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carle methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other astropy classes, such as
\astropyskycoord\ and \astropyTime.

\paragraph{Time series.}  From sampling a continuous variable at fixed
times to counting events binned into time windows, many different
areas of astrophysics require the manipulation of time series.  These
are often logically tables, and the new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to tables of data as a
function of time, where the data can either represent samples or
averages over particular time bins.  The new classes offer a number of
special methods to manipulate time series (folding, resampling, etc.)
and to read different data formats (such as {\em Kepler} lightcurves,
etc.).

Also part of \astropysubpkg{timeseries} are common analysis routines,
including Lomb-Scargle and box-least-squares periodogrammes.

\paragraph{Spectral Coordinates.} Measurements are often taken at
specific ``spectral coordinates'', be they frequencies, wavelengths,
or photon energies.  The new \astropySpectralCoord\ class allows to
transform between these as well as to equivalent Doppler velocities.
It builds on the \astropyQuantity, which can also represent these
kinds of physical quantities, and allows conversion via dedicated
equivalencies, but aims to make this more straightforward.
Furthermore, it can be made aware of the observer and target reference
frames, allowing for instance transformation from telescope-centric (or
topocentric) frames to Barycentric or Local Standard of Rest
(LSRK and LSRD) velocity frames.

\subsubsection*{Interoperability} \label{sec:core-features-interoperability}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}

@@

Summarize funding sources (Moore, NASA) and amounts and what this has been used
for.

@@

\secauthor{Lía Corrales, David Shupe, Kelle Cruz + Learn team}

\subsubsection{Current status and scope}

\textit{Learn Astropy} is an umbrella term that acknowledges the broad
educational efforts made by the Astropy Project, which are led by the Learn
Team.
The efforts focus on developing online content and workshops covering
astronomy-specific coding tasks in Python.
As introduced in \citet{astropy:2018}, there are four different types of
Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
lessons that are published in HTML format online; \textit{guides}, which are a
series of lessons providing a foundational resource for performing certain
type of astronomical analyses; \textit{examples}, which are snippets of code
that showcase a short task that can be performed with Astropy or an affiliated
package; and \textit{documentation}, which is contained within the code base.
This categorization drives content development, infrastructure choices, and the
appearance of the Learn Astropy website.
The Learn Team meets weekly to work on creating, expanding, improving these
educational resources.

\begin{description}

\item[Learn Astropy website] The Learn Team re-launched the main website and
search interface for Learn Astropy in 2021 with a new infrastructure platform,
built around full-text search and interactive filtering functionality, with the
goal of making content discoverable as the Learn Astropy content catalog
expands.
This work has been supported in part by a grant from the Dunlap Institute.
We have adopted Algolia, a search-as-a-service cloud platform, to store the
full-text and metadata records of Learn Astropy's content.
The new Learn Astropy website is a JavaScript (Gatsby/React) application that
uses the Algolia service to power its search and filtering user interface.
Our Python-based application, Learn Astropy Librarian, populates data into the
Algolia service. We tuned the Librarian around specific content formats (such as
Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
accurately index content and heuristically extract metadata.
A consequence of the new platform is that we now maintain and compile content
separately from the website application itself, enabling new content types.
Tutorials, which are written as Jupyter Notebooks, are now compiled into their
own Learn Astropy sub-site using
\href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
infrastructure, are also deployed as separate websites using GitHub pages.
This architecture opens future possibilities of indexing third-party content,
hosted elsewhere, such as on institutional websites.

\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.

\item[Guides] {\it ccdproc} Jupyter book has been completed.

\item[Workshops] The Project has been conducting workshops at winter
meetings of the American Astronomical Society since AAS 225 in January 2015.
Up to the start of the coronavirus pandemic, these were full-day in-person
workshops with as many as 90 participants and a dozen facilitators from the
project.
During the pandemic, these workshops were moved to an online format and split
into basic and advanced sessions.
Additionally, beginning with the AAS 238 online meeting, the workshops have been
expanded to Summer AAS meetings.
The Learn team finds that the workshop audience is best found as AAS meetings as
opposed to more general Python meetings.

The Astropy Project provided another mode of community engagement at AAS
Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
AAS Exhibit Hall.
Funding for the exhibit hall was provided alternately by NumFOCUS and later by
the Moore Foundation funding (\textbf{check this}).
The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
during the virtual AAS 237 meetings, to provide the general astronomy community
information and access to experts on a variety of open source astronomical
tools.
\end{description}

\subsubsection{Learn vision for the future}

The Learn plan going forward is to continue to improve the tutorial website and
facilate new content;
to index third-party tutorials; and to look for supportable opportunities to
expand the reach of Astropy workshops beyond the American Astronomical Society
meetings.

@@

\section{Community Engagement}

\begin{itemize}
\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community. This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace. The Moore Foundation grant allows this Slack space to
be on a paid plan; additionally NumFOCUS has negotiated a special rate for
open-source projects. While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source. The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
a Discourse site which is more open than the Facebook group and more user-oriented
than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
developers are able to invoice as independent contractors the time they
spend helping users on these forums.
\end{itemize}

>>>>

Summarize major pieces of development since the last paper (since v2.0). A few
ideas (taken from What's New pages) below, but feel free to expand on or remove
from this list:
\begin{itemize}
    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
    \item TODO: something about WCS, SpectralCoord?
\end{itemize}

@@

% Summarize funding sources (Moore, NASA) and amounts and what this has been
% used for.

With the goal of establishing a sustainable financial model for the Astropy
Project, the Coordination Committee under the new governance charter (see \S
\ref{sec:project-governance}) established a standing Finance committee.
This committee oversee the planning and allocation of finances on behalf
of the project. The Astropy Project accepts funds from institutions as well as
individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
supports and promotes world-class, innovative, open source scientific computing.}.
Previously, no direct financial support was available for the project development,
and NUMFOCUS covered most of the incurred operational costs. However, transitioning
to long-term financial stability meant having an influx of funds in the form of
institutional support, and hence the Astropy Project applied for and successfully
acquired four grants. These grants are administered by NumFOCUS on behalf of the
project, providing a ``neutral'' space not tied to any particular astronomy
institution. The two major ones are briefly summarized below.

\begin{itemize}
    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project". Over a three-year term, this grant supported
    sustenance of the project by providing funds for (1) the development and
    maintenance of the \astropypkg and its infrastructure to existing project
    members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the formalization
    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
    improvement of equity, diversity, and inclusion efforts in the project (see \S
    \ref{sec:project-ide}). Some of the money supported travel for meetings,
    conferences, and workshops. This generous grant helped reduce reliance on
    volunteer-driven maintenance of the openly-developed package and paved
    the way to acquire funding from federal agencies: a big milestone for the
    success of the project.

    \item NASA: A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages that need support, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{itemize}

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All these successful proposals show that the Astropy project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

% ADD text for the new finance APE once it has a doi

@@

\secauthor{Lía Corrales, David Shupe + Learn team}

\begin{itemize}
\item Current status and scope
\item Vision for the future (review what we said in v2.0 paper)
\item User forums and engagement with user base
\item Workshops
\end{itemize}

@@

\secunfilled
